### 2021-09-29 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betadiversity_response@d5c2f7fb602e23a83bab8f09321a9a8fc7286211 🚀

The approach to quantifying the difference between pairs of networks
established in Poisot \emph{et al.} (2012) is a simple extension of the
overall method by Koleff \emph{et al.} (2003) for species dissimilarity
based on presence-absence data. The objects to compare, \(X_1\) and
\(X_2\), are partitioned into three values, \(a = |X_1 \cup X_2|\),
\(b = |X_2 \setminus X_1|\), and \(c = |X_1 \setminus X_2|\), where
\(|\dot|\) is the cardinality of set \(\dot\) (the number of elements it
contains), and \(\setminus\) is the set substraction operation. In the
perspective of species composition comparison, \(X_1\) and \(X_2\) are
the sets of species in either community, so that if
\(X_1 = \{x, y, z\}\) and \(X_2 = \{v, w, x, y\}\), we have
\(X_1 \cup X_2 = \{v, w, x, y, z\}\), \(X_1 \cap X_2 = \{x, y\}\),
\(X_2 \setminus X_1 = \{v, w\}\), and \(X_1 \setminus X_2 = \{z\}\). The
core message of Koleff \emph{et al.} (2003) is that the overwheling
majority of measures of \(\beta\)-diversity can be re-expressed as
functions that operate on the cardinality of these sets -- this allows
to focus on the number of unique and common elements, as outlined in
fig.~\ref{fig:conceptual}.

@@

Applying this framework to networks requires a few additional
definitions. Although ecologists tend to think of networks as their
adjacency matrix (as is presented in fig.~\ref{fig:conceptual}), this
representation is not optimal to reach a robust understanding of which
elements should be counted as part of which set when measuring network
dissimilarity. For this reason, we need fall back on the definition of a
graph as a pair of sets, wherein \(\mathcal{G} = (V, E)\). These two
components \(V\) and \(E\) represent vertices (nodes, species) and edges
(interactions), where \(V\) is specifically a set containing the
vertices \(\mathcal{G}\), and \(E\) is a set of ordered pairs, in which
every pair is composed of two elements of \(V\); an element \(\{i,j\}\)
in \(E\) indicates that there is an interaction \emph{from} species
\(i\) to species \(j\) in the network \(\mathcal{G}\).

@@

\[\mathcal{M} = (V_c \cup V_n, E_c \cup E_{sn} \cup E_{un}) \,,\]

where \(V_c\) is the set of shared species, \(V_k\) are the species
belonging only to network \(k\), \(E_c\) are the shared edges, and
\(E_{sk}\) and \(E_{uk}\) are the interactions unique to \(k\)
involving, respectively, only species in \(V_c\), and at least one
species from \(V_k\).

@@

The decomposition of \(\beta\)-diversity from Poisot \emph{et al.}
(2012) uses these components to measure \(\beta_{os}\) (the interaction
dissimilarity between shared species, which Fründ (2021) terms
``rewiring''), and \(\beta_{wn}\) (the overall dissimilarity including
non-shared species). We can express the components \(a\), \(b\), and
\(c\) of Koleff \emph{et al.} (2003) as the cardinality of the following
sets:

@@

These decompositions are used to perform the calculations of
\(\beta\)-diversity in the \texttt{EcologicalNetworks.jl} package
(Banville \emph{et al.} 2021) for Julia, which I use for the following
numerical experiments.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Banville2021ManJl}{}%
Banville, F., Vissault, S. \& Poisot, T. (2021). Mangal.jl and
EcologicalNetworks.jl: Two complementary packages for analyzing
ecological networks in Julia. \emph{Journal of Open Source Software}, 6,
2721.

\leavevmode\hypertarget{ref-Baronio2021NatFir}{}%
Baronio, G.J., Souza, C.S., Maruyama, P.K., Raizer, J., Sigrist, M.R. \&
Aoki, C. (2021). Natural fire does not affect the structure and beta
diversity of plant-pollinator networks, but diminishes floral-visitor
specialization in Cerrado. \emph{Flora}, 281, 151869.

>>>>

The approach to quantifying the difference between pairs of networks
established in Poisot \emph{et al.} (2012) is a simple extension of the
overall method by Koleff \emph{et al.} (2003) for species dissimilarity
based on presence-absence data. The objects to compare, \(X_1\) and
\(X_2\), are partitioned into three values, \(a = |X_1 \cup X_2|\),
\(b = |X_2 \setminus X_1|\), and \(c = |X_1 \setminus X_2|\), where
\(|\cdot|\) is the cardinality of set \(\cdot\) (the number of elements
it contains), and \(\setminus\) is the set substraction operation. In
the perspective of species composition comparison, \(X_1\) and \(X_2\)
are the sets of species in either community, so that if
\(X_1 = \{x, y, z\}\) and \(X_2 = \{v, w, x, y\}\), we have
\(X_1 \cup X_2 = \{v, w, x, y, z\}\), \(X_1 \cap X_2 = \{x, y\}\),
\(X_2 \setminus X_1 = \{v, w\}\), and \(X_1 \setminus X_2 = \{z\}\). The
core message of Koleff \emph{et al.} (2003) is that the overwheling
majority of measures of \(\beta\)-diversity can be re-expressed as
functions that operate on the cardinality of these sets -- this allows
to focus on the number of unique and common elements, as outlined in
fig.~\ref{fig:conceptual}.

@@

Applying this framework to networks requires a few additional
definitions. Although ecologists tend to think of networks as their
adjacency matrix (as is presented in fig.~\ref{fig:conceptual}), this
representation is not optimal to reach a robust understanding of which
elements should be counted as part of which set when measuring network
dissimilarity. For this reason, we need fall back on the definition of a
graph as a pair of sets, wherein \(\mathcal{G} = (V, E)\). These two
components \(V\) and \(E\) represent vertices (nodes, species) and edges
(interactions), where \(V\) is specifically a set containing the
vertices of \(\mathcal{G}\), and \(E\) is a set of ordered pairs, in
which every pair is composed of two elements of \(V\); an element
\(\{i,j\}\) in \(E\) indicates that there is an interaction \emph{from}
species \(i\) to species \(j\) in the network \(\mathcal{G}\). The
adjancency matrix \(\mathbf{A}\) of this network would therefore have a
non-zero entry at \(A_{ij}\).

@@

\[\mathcal{N} = (V_c \cup V_n, E_c \cup E_{sn} \cup E_{un}) \,,\]

where \(V_c\) is the set of common species, \(V_m\) and \(V_n\) are the
species belonging only to network \(m\) and \(n\) (respectivly), \(E_c\)
are the common edges, and \(E_{sm}\) and \(E_{um}\) are the interactions
unique to \(k\) involving, respectively, only species in \(V_c\), and at
least one species from \(V_m\) (the same notation applies for the
subscript \(_{n}\)).

@@

The decomposition of \(\beta\)-diversity from Poisot \emph{et al.}
(2012) uses these components to measure \(\beta_{os}\) (``rewiring''),
and \(\beta_{wn}\) (the overall dissimilarity including non-shared
species). We can express the components \(a\), \(b\), and \(c\) of
Koleff \emph{et al.} (2003) as the cardinality of the following sets:

@@

It is fundamental to note that these components can be measured entirely
from the interactions, and that the number of species in either network
are never directly involved.

In the following sections, I present a series of calculations aimed at
expressing the values of \(\beta_{os}\), \(\beta_{wn}\), and therefore
\(\beta_{st}\) as a function of species sharing probability (as a proxy
for mechanisms generating turnover), and link rewiring probability (as a
proxy for mechanisms generating differences in interactions among shared
species). These calculations are done using \texttt{Symbolics.jl}
(\textbf{Gowda2021HigSym?}), and subsequently transformed in executable
code for \emph{Julia} (\textbf{Bezanson2017JulFre?}), used to produce
the figures.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Baronio2021NatFir}{}%
Baronio, G.J., Souza, C.S., Maruyama, P.K., Raizer, J., Sigrist, M.R. \&
Aoki, C. (2021). Natural fire does not affect the structure and beta
diversity of plant-pollinator networks, but diminishes floral-visitor
specialization in Cerrado. \emph{Flora}, 281, 151869.
